Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-14 talk/dls2012/paper.tex
copyright data
add remaining scimark lua benchmarks to runner

\conferenceinfo{IWTC '11}{XXX} 
\copyrightyear{2011} 
\copyrightdata{[to be supplied]}

@@

The work described in this paper was done in the context of the PyPy
project.\footnote{\texttt{http://pypy.org}} PyPy is a framework for implementing
dynamic languages efficiently~\cite{armin_rigo_pypys_2006}. When implementing a
language with PyPy, one writes an interpreter for the language in RPython~\cite{davide_ancona_rpython:_2007}.
RPython (``Restricted Python``) is a subset
of Python chosen in such a way that it can be efficiently translated to a
C-based VM by performing type inference.

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs to be executed the first time and then the result
can be reused for all other appearances. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}

@@

Other interesting interpreters that are helped greatly by this optimization are
for example our Prolog interpreter written in
RPython~\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain
tight
loops that perform list processing. Furthermore we experimented with a Python library
for writing numerical kernels doing array manipulation. The exact extent is
out of scope for this paper.

@@

Loop invariant code motion optimizations are a well-known approach to optimize
loops~\cite{muchnick_advanced_1997}. Therefore, the effects that the
optimizations described here achieve are not in any way new. However, we think
that achieving them in the way described in this paper is simpler than writing
explicit algorithms.
\cfbolz{more explicit listing of prior work goes here}

@@

The current approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees~\cite{andreas_gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.

>>>>

\conferenceinfo{DLS'12,} {October 22, 2012, Tucson, Arizona, USA.}
\CopyrightYear{2012}
\copyrightdata{978-1-4503-1564-7/12/10}

@@

The work described in this paper was done in the context of the PyPy
project.\footnote{\texttt{http://pypy.org}} PyPy is a framework for implementing
dynamic languages efficiently~\cite{rigo_pypys_2006}. When implementing a
language with PyPy, one writes an interpreter for the language in RPython~\cite{ancona_rpython:_2007}.
RPython (``Restricted Python``) is a subset
of Python chosen in such a way that it can be efficiently translated to a
C-based VM by performing type inference.

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs to be executed the first time and then the result
can be reused for all other appearances. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{diwan_type-based_1998}. In addition writes
on newly allocated objects can never change the value of old existing ones.}

@@

Other interesting interpreters that are helped greatly by this optimization are
for example our Prolog interpreter written in
RPython~\cite{bolz_towards_2010}. Prolog programs often contain
tight
loops that perform list processing. Furthermore we experimented with a Python library
for writing numerical kernels doing array manipulation. The exact extent is
out of scope for this paper.

@@

Loop invariant code motion optimizations are a well-known approach to optimize
loops~\cite{muchnick_advanced_1997}. Therefore, the effects that the
optimizations described here achieve are not in any way new. However, we think
that achieving them in the way described in this paper is simpler than writing
explicit algorithms.

Loop invariant code motion has been part of early compilers in the 1960s and
1970s~\cite{allen_catalogue_1971}. The approach for achieving loop invariant
code motion is typically to perform partial redundancy elimination. The
approach was first proposed by Morel and Renvoise~\cite{morel_global_1979}. It
involves solving data flow problems usually involding bidirection data flow
equations. After improvements~\cite{chow_portable_1984,
dhamdhere_practical_1991} this approach was followed by the work of Knoop
et.al.~\cite{knoop_lazy_1992} who cleany separated the problem into a backward
and forward data flow analysis. Implementing partial redundancy elimination in
compilers that use SSA form \cite{chow_new_1997} simplified the algorithms
because no iterative data flow analysis is needed any more.

@@

The current approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees~\cite{gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.
